---
title: "SORT-seq dataset with velocity preparations"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

## *Setting Parameters:* 
Make sure all the metadata is available in the smples.tsv files from s2s, those columns will be loaded into the cellseq object
```{r}
## Parameters for processing dataset and metadata ##
#barcode_file = "/scratch/snabel/scrna/barcode_384.tab"
barcode_file = '/ceph/rimlsfnwi/data/moldevbio/zhou/jsmits/scRNAseq_iPSC_iLSCs/genomes/spike_in/barcode_384.tab'
barcode_file_2 = "/ceph/rimlsfnwi/data/moldevbio/zhou/jsmits/scRNAseq/barcode_2_well.tab"
#MT_genes_file = "/scratch/snabel/scrna/genomes/genome_addons/MT_genes.txt"
MT_genes_file = '/ceph/rimlsfnwi/data/moldevbio/zhou/jsmits/scRNAseq_iPSC_iLSCs/genomes/spike_in/MT_genes.txt'
s2s_results = "/ceph/rimlsfnwi/data/moldevbio/zhou/jsmits/scRNAseq_iPSC_iLSCs/results"
s2s_sample_file = "/ceph/rimlsfnwi/data/moldevbio/zhou/jsmits/scRNAseq_iPSC_iLSCs/samples.tsv"

#read s2s sample file:
sample_info <- read.table(s2s_sample_file, header = TRUE, sep = '\t',comment.char = "#")

lab_col = 'cell_type'

#subset data
subset_id <- 'cell_type'
#filtering = c('ESC','d11_iLSC','d24_iLSC', 'LSC1','LSC2')
explore_violin <- c("ABCG2","TP63","LGR5","ABCB5","BMI1","FZD7","CEBPD","CD200","CD109","POU5F1","PAX6","KRT3","KRT12","KRT14","KRT15","MKI67","CDKN1B")
marker_gene_file <- '/ceph/rimlsfnwi/data/moldevbio/zhou/jsmits/scRNAseq_iPSC_iLSCs/LSC_Marker_Genes.csv'

## Filtering of the dataset ##
# Settings for genes
gene_tresh = 1
amount_cells_expr = 1
# Settings for cells
total_counts_tresh = 2000
total_feat_tresh = 1000
ERCC_pct_max <- 20
mt_pct_max <- 40

## Seurat Normalization, HVG selection (vst) & Scaling (and Regression) ##
nHVG = 2000
# Regression performed on the following variables:
vars_to_regress = c("nCount_sf", "nFeature_sf",'pct_counts_MT') # If no regression desired: NULL
vars_cell_cycle = c("S.Score","G2M.Score")

## Dimensionality reduction ##
# For PCA to run on
pcs_max = 70 

## Storing results ##
workdir <- "/ceph/rimlsfnwi/data/moldevbio/zhou/jsmits/scRNAseq_iPSC_iLSCs/analysis/"
# if regression is performed: this will already be included in the folder name
result_descript = "_results_PreprocessingDataset_complete_clean"

## Location of scripts ##
source("/ceph/rimlsfnwi/data/moldevbio/zhou/jsmits/tools/scripts/qc_ercc_384plot.R")
source("/ceph/rimlsfnwi/data/moldevbio/zhou/jsmits/tools/scripts/qc_umis_384plot.R")
source("/ceph/rimlsfnwi/data/moldevbio/zhou/jsmits/tools/scripts/read_kb_dataset_s2s.R")

dateoftoday <- gsub("-", "", as.character(Sys.Date()))
resultsdir <- paste0(workdir, dateoftoday)
system(paste("mkdir -p ", resultsdir))
knitr::opts_knit$set(root.dir = normalizePath(resultsdir))
knitr::opts_knit$set(root.dir = normalizePath(workdir))

#knitr::opts_knit$set(root.dir = "/ceph/rimlsfnwi/data/moldevbio/zhou/jsmits/scRNAseq_iPSC_iLSCs/analysis")

#run first time (no good anaconda packages, sadly):
#BiocManager::install("DO.db", force = TRUE)#update none
#BiocManager::install("R.utils", force = TRUE)#update none
#remotes::install_github('satijalab/seurat-wrappers') #update none


# Loading the important repositories # 
require("devtools")
library('ggplot2')
library('plyr')
library('dplyr')
library('tidyr')
library('mvoutlier')
library('limma')
library('knitr')
library('SingleCellExperiment')

library("scater")
library("Seurat")
library("scran")
library("RColorBrewer")
library("plot3D")
library("stringr")
library("SAVER")
library("parallel")
library('progeny')
library("ComplexHeatmap") 
library("org.Hs.eg.db")
library("clusterProfiler") 
library("ggpubr")
library("circlize")
library("cowplot")
library('clustree')
library('tidyverse')

library('monocle')
library('SeuratWrappers')


```

### Creating the SingleCellExperiment object
The counts table is loaded along with the metadata of the cells within an Scater usable object. Scater will be used to look into the quality of the data and to help with filtering out bad cells or genes.
```{r loading dataset, echo = TRUE, warning = FALSE}
## Splice seperated dataset:
spliced.data.df = read_kb_counts(paste(s2s_results,"/kallistobus/", sep = ""), "spliced", barcode_file = barcode_file)
unspliced.data.df = read_kb_counts(paste(s2s_results,"/kallistobus/", sep = ""), "unspliced", barcode_file = barcode_file)
```


#calculate spliced/unspliced percentage
```{r}
# Percentage of reads unspliced
print(paste0('percentage unspliced reads = ', (sum(unspliced.data.df)/(sum(spliced.data.df)+sum(unspliced.data.df))
)*100))
all_cells <- intersect(colnames(spliced.data.df),colnames(unspliced.data.df))
unspliced.data.df <- unspliced.data.df[,all_cells]
spliced.data.df <- spliced.data.df[,all_cells]
#identical(colnames(spliced.data.df),colnames(unspliced.data.df))

# The default data.df will be the spliced dataset (shorter to type)
data.df <- spliced.data.df
```
#load sample metadata from the seq2science file, find in which well each fastq file was located
```{r phenotable}
## Setting up the phenotable ##
phenodata <- data.frame(row.names=colnames(data.df))
phenodata$names <- row.names(phenodata)

#add well ID as metadata, get the well info from the last _* regex from each column name
well_id <- function(x){
  x <- str_split(x, '_')
  n_sub_strings <- length(x[[1]])
  x <- x[[1]][n_sub_strings]
}
phenodata$well <- sapply(phenodata$names, function(x) well_id(x))

#generate the original sample name, by removing the well info, merge with the s2s metadata
phenodata$name_nowell <- mapply(function(x,y)gsub(y,"",x),phenodata$names,paste('_',phenodata$well,sep=""))
phenodata$sample <- gsub("GRCh38_ERCC_reporter_", "",phenodata$name_nowell)
phenodata$barcode_name <- paste(phenodata$name_nowell,phenodata$well, sep = "")
phenodata <- merge(phenodata,sample_info, by.x = 'sample', by.y = 'sample')

# Only take the entries that are matchable with the counttable entries:
pheno_matched <- phenodata[phenodata$names %in% colnames(data.df),]
# Matching phenodata with the dataset ordering
pheno_ordered <- pheno_matched[match(colnames(data.df),pheno_matched$names),]
identical(pheno_ordered$names, colnames(spliced.data.df))
write.csv(phenodata, paste0(resultsdir,"/phenodata_kbordered.csv"))
```

Subset to only use the relevant samples
```{r}
#pheno_ordered <- pheno_ordered[pheno_ordered[[subset_id]] %in% filtering,]
#pheno_ordered$cell_type <- revalue(pheno_ordered$cell_type, c("ESC"="d0_iLSC"))
#pheno_ordered
data.df <- data.df[,colnames(data.df) %in% pheno_ordered$names]
```

## Plate Overview QC
```{r}
# Make a list of cell-names compatable with the excel file: plate#_A1, plate#_A2 etc.
plate_order <- read.table(barcode_file_2, sep = "\t", col.names = c("well","barcode"))
# Make a vector with all plate numbers
platenrs <- unique(phenodata$name_nowell)
pdf(paste0(resultsdir,"/1.QC_per_Plate.pdf"), paper = "USr")
# settings for the plate diagnostics pdf 
par(mfrow=c(2,2), mar = c(5,4,4,2) + 0.1, cex.main = 1)
# Iterate over all plates, order cells in the order of visualization
for (plate in platenrs){
  # use the order of cells from te barcode file (this is A1, A2, A3, etc to P24)
  primer_order <- paste(plate, plate_order$well, sep="")
  # if wells are missing on the plate, these need to be added and given a value of 0
  missing_wells <- primer_order[!primer_order %in% colnames(spliced.data.df)]
  cols_to_add <- data.frame(matrix(ncol = length(missing_wells), nrow = length(rownames(spliced.data.df))))
  colnames(cols_to_add) <- missing_wells
  cols_to_add[is.na(cols_to_add)] <- 0
  diag_plate <- cbind(spliced.data.df[,grep(plate, colnames(spliced.data.df))], cols_to_add)
  # phenodata contains same cellid entry + rowname as used in dataset
  cells_order <- colnames(diag_plate[,match(primer_order, colnames(diag_plate))])
  
  # match dataset cells order with wells in the visualization
  tmp <- as.matrix(diag_plate[,cells_order])
  QC_umis_384plot(tmp, paste(plate, "UMI_QC", sep = "_"))
  QC_ERCC_384plot(tmp[grep("^ERCC", rownames(diag_plate)),], paste(plate, "ERCC_QC", sep = "_"))
  rm(tmp)
}
dev.off()

```


#### Creating a SingleCellExperiment object
```{r build SCE}
count_matrix <- as.matrix(data.df)
sce <- SingleCellExperiment(assays = list(counts = count_matrix), colData = pheno_ordered, rowData = rownames(count_matrix))
MT_genes <- read.table(MT_genes_file)[,1]
# Adding spike-in information:
is.spike <- grepl("^ERCC", rownames(sce))
sce <- splitAltExps(sce, ifelse(is.spike, "ERCC", "gene"))
is.spike <- rownames(sce) %in% MT_genes
sce <- splitAltExps(sce, ifelse(is.spike, "MT", "gene"))
# Calculate the quality metrics:
sce <- addPerCellQC(sce)

pdf(paste0(resultsdir,'/2a.QC_counts_histograms.pdf') ,width=10,height=10,paper='special') 
hist(sce$sum, breaks = 100)
abline(v = total_counts_tresh, col = "red")
# Looking at the amount of unique genes per sample
# This is the amount with ERCC included.
hist(sce$detected, breaks = 100)
abline(v= total_feat_tresh, col = "red")
dev.off()

```

```{r}
nmads = 1.5
#sce <- na.omit(sce)
libsize.drop <- isOutlier(as.numeric(sce$sum), nmads=nmads, type="lower", log=TRUE)
feature.drop <- isOutlier(as.numeric(sce$detected), nmads=nmads, type="both", log=TRUE)
below_feature_thresh <- sce$detected < total_feat_tresh
MT.drop <- isOutlier(sce$altexps_MT_percent, nmads=nmads, type="higher")
spike.drop <- isOutlier(sce$altexps_ERCC_percent, nmads=nmads, type="higher")
#sce$keep <- !(libsize.drop| feature.drop | spike.drop | MT.drop | below_feature_thresh)
sce_qc <- sce[,!(libsize.drop| feature.drop | spike.drop | MT.drop | below_feature_thresh)]
mean(table(sce_qc$sample))


```

#------------------------
## Filtering the genes ##
#------------------------

```{r filter genes}
#---------------------------

# Create the quality-checked dataset:
write.table("spliced_qc_counts.tsv", col.names = NA, quote = FALSE)
#lets see how much each variable explains variability of the log gene scores
sce_qc <- logNormCounts(sce_qc) 
vars <- getVarianceExplained(sce_qc, 
    variables=c("timepoint", "cell_type", "batch", "sum"))

# https://www.bioconductor.org/packages/devel/bioc/vignettes/scater/inst/doc/vignette-qc.html#identifying-outliers-on-all-qc-metrics
pdf(paste0(resultsdir,'/4.QC_genes_expressed.pdf') ,width=10,height=5,paper='special') 
plotColData(sce_qc, x = "sum", y="detected", colour_by="timepoint") 
plotColData(sce_qc, x = "sum", y="altexps_MT_percent", 
    other_fields="timepoint") + facet_wrap(~timepoint)
plotExplanatoryVariables(vars)
# The reads consumed by the top 50 expressed genes:
plotHighestExprs(sce_qc)
dev.off()
```

## Build unspliced assay
Select the same cells and genes as in the spliced dataset
```{r build SCE 2}
cells_use <- colnames(sce_qc)
genes_use <- rownames(sce_qc)

sce_us <- SingleCellExperiment(assays = list(counts = as.matrix(unspliced.data.df)), colData = pheno_matched, rowData = rownames(unspliced.data.df))
# Adding spike-in information:
is.spike <- grepl("^ERCC", rownames(sce_us))
sce_us <- splitAltExps(sce_us, ifelse(is.spike, "ERCC", "gene"))
is.spike <- rownames(sce_us) %in% MT_genes
sce_us <- splitAltExps(sce_us, ifelse(is.spike, "MT", "gene"))
# Calculate the quality metrics:
sce_us <- addPerCellQC(sce_us)

# Dataset after filtering:
sce_usmatch <- sce_us[genes_use,cells_use]

# Arbitrary thresholds:
# Looking at the total number of RNA molecules per sample
# UMI counts were used for this experiment
pdf(paste(resultsdir,'2a.QC_unspliced_reads.pdf',sep="/") ,width=20,height=20,paper='special') 
par(mfrow=c(2,2))
hist(sce_us$sum, breaks = 100)
hist(sce_us$detected, breaks = 100)
hist(sce_usmatch$sum, breaks = 100)
hist(sce_usmatch$detected, breaks = 100)
dev.off()
```

## make a pseudobulk table:
```{r}
pseudobulk_df <- as.data.frame(row.names(counts(sce_qc)))
for (sample in unique(sce_qc$sample)){
  pseudobulk_df[[sample]] <- rowSums(counts(sce_qc)[,sce_qc$sample == sample])
}
row.names(pseudobulk_df) <- pseudobulk_df$`row.names(counts(sce_qc))`
pseudobulk_df$`row.names(counts(sce_qc))` <- NULL
write.table(pseudobulk_df, file = paste0(resultsdir,'/pseudobulk.tsv'), sep = '\t')
```

## Build Seurat object with unspliced and spliced assay
## make a filtered and unfiltered object to show the difference between before and after filtering
```{r create seurat object}
metadata_2_include <- c(colnames(phenodata),"altexps_ERCC_percent","altexps_MT_percent")
seur_obj <- CreateSeuratObject(counts = counts(sce_qc), assay = "sf", meta.data = as.data.frame(colData(sce_qc)[,colnames(colData(sce_qc)) %in% metadata_2_include]))
#seur_obj <- CreateSeuratObject(counts = counts(sce_endo), assay = "counts", meta.data = as.data.frame(colData(sce_endo)[,colnames(colData(sce_endo)) %in% metadata_2_include]))

seur_obj_all <- CreateSeuratObject(counts = counts(sce), assay = "sf", meta.data = as.data.frame(colData(sce)[,colnames(colData(sce)) %in% metadata_2_include]))
unspliced_match <- unspliced.data.df[genes_use,cells_use]
unspliced_match <- as.matrix(unspliced_match)
seur_obj[["uf"]] <- CreateAssayObject(counts = unspliced_match)


#set the timepoints in the correct order
cluster_order <- c('d0','d11','d10','d24','mixABCG2','purABCG2','adult_LSC')
seur_obj$timepoint <- factor(x = seur_obj$timepoint, levels = cluster_order)

```

## RUN SAVER on the good quality cells to improve coverage of low expressed genes
## We will use this imputated data only for gene expression plots, not for clustering/dimensionality reduction.
Run SAVER on the countmatrix, save the imputed counts to the 'data.saver' object within the Seurat object
```{r message=FALSE, warning=FALSE}
# nmads_file_loc <- '/ceph/rimlsfnwi/data/moldevbio/zhou/jsmits/scRNAseq_iPSC_iLSCs/analysis/data.saver_nmads_stringent.rds'
# 
# if (file.exists(nmads_file_loc)){print('SAVER file exists, loading and using it')
#   data.saver <- readRDS(nmads_file_loc)
# } else {print('no Saver file found, generating one')
# data.saver <- saver(seur_obj[['sf']]@counts, ncores = 4)
# saveRDS(data.saver, file = nmads_file_loc)
# }
# seur_obj[['sf_SAVER']] <- CreateAssayObject(data = data.saver$estimate, min.cells = 1, min.features = 100)

```

```{r}
library(gridExtra)

Idents(seur_obj) <- "sample"
Idents(seur_obj_all) <- "sample"
df <- as.data.frame(table(seur_obj$descriptive_name))
colnames(df) <- c('descriptive_name', 'cells after QC')

pdf(paste(resultsdir,'3.QC_depth_ERCC_MT.pdf',sep="/") ,width=20,height=12,paper='special')
VlnPlot(object = seur_obj_all, features = ("nCount_sf"), assay = 'sf') + scale_y_continuous(limits = c(0,20000)) + ggtitle('total UMI counts all cells') 
VlnPlot(object = seur_obj, features = c("nCount_sf"), assay = 'sf') + scale_y_continuous(limits = c(0,20000)) + ggtitle('total UMI counts filtered cells') 
VlnPlot(object = seur_obj_all, features = c("nFeature_sf"), assay = 'sf') + scale_y_continuous(limits = c(0,5000)) + ggtitle('genes measured all cells') 
VlnPlot(object = seur_obj, features = c("nFeature_sf"), assay = 'sf')  + scale_y_continuous(limits = c(0,5000)) + ggtitle('genes measured filtered cells') 
VlnPlot(object = seur_obj_all, features = c("altexps_MT_percent"))+ scale_y_continuous(limits = c(0,40)) + ggtitle('% mitochondrial reads all cells') 
VlnPlot(object = seur_obj, features = c("altexps_MT_percent"))+ scale_y_continuous(limits = c(0,40)) + ggtitle('% mitochondrial reads filtered cells') 

VlnPlot(object = seur_obj_all, features = c("altexps_ERCC_percent"))+ scale_y_continuous(limits = c(0,10)) + ggtitle('% ERCC reads all cells') 
VlnPlot(object = seur_obj, features = c("altexps_ERCC_percent"))+ scale_y_continuous(limits = c(0,10)) + ggtitle('% ERCC reads filtered cells')

FeatureScatter(seur_obj_all, feature1 = "nCount_sf", feature2 = "nFeature_sf") + ggtitle('% counts vs genes measured all cells') 
FeatureScatter(seur_obj, feature1 = "nCount_sf", feature2 = "nFeature_sf") + ggtitle('% counts vs genes measured filtered cells') 
plot.new()
grid.table(df)
dev.off()

```

#Perform cell-cycle scoring
```{r}
seur_obj <- NormalizeData(
    object = seur_obj, assay = "sf",
    normalization.method = "LogNormalize", 
    scale.factor = 10000
)

seur_obj <- NormalizeData(
    object = seur_obj, assay = "uf",
    normalization.method = "LogNormalize", 
    scale.factor = 10000
)
seur_obj <- FindVariableFeatures(seur_obj, selection.method = "vst")
seur_obj <- ScaleData(seur_obj, features = rownames(seur_obj), assay = 'sf', vars_to_regress = c('altexps_MT_percent', 'nCount_sf','nFeature_sf', 'altexps_ERCC_percent' ))
seur_obj <- CellCycleScoring(seur_obj, s.features = cc.genes.updated.2019$s.genes, g2m.features = cc.genes.updated.2019$g2m.genes, set.ident = TRUE)
seur_obj <- RunPCA(seur_obj, features = c(cc.genes.updated.2019$s.genes, cc.genes.updated.2019$g2m.genes))
pdf(paste(resultsdir,'5.cell_cycle_markers.pdf',sep="/") ,width=12,height=6,paper='special')
Idents(seur_obj) <- "Phase"
PCAPlot(seur_obj)
Idents(seur_obj) <- "timepoint"
PCAPlot(seur_obj)

RidgePlot(seur_obj, features = c("PCNA", "TOP2A", "MCM6", "MKI67"), assay= 'sf', ncol = 2, slot = "data")
RidgePlot(seur_obj, features = c("S.Score", "G2M.Score"), ncol = 2, group.by = "timepoint")
dev.off()
```
Vizualize the PCAs
```{r}
library(grid)
library(gridExtra)

seur_obj <- RunPCA(seur_obj, verbose = FALSE, npcs = 50)
mat <- Seurat::GetAssayData(seur_obj, assay = "sf", slot = "scale.data")
pca <- seur_obj[["pca"]]

# Get the total variance:
total_variance <- sum(matrixStats::rowVars(mat))

eigValues = (pca@stdev)^2  ## EigenValues
varExplained = eigValues / total_variance
Stdev(object = seur_obj[["pca"]])[1]


Idents(seur_obj) <- "timepoint"
  pdf(paste0(resultsdir,'/6.Principle_components.pdf') ,width=12,height=6,paper='special')
  print(ElbowPlot(seur_obj))
  for (pcs in c(1:10)){
    pc1_viz <- pcs*2-1
    pc2_viz <- pcs*2
    y_label = paste0(paste0(paste0("PC",pc2_viz), ' stdev:  '),round(Stdev(object = seur_obj[["pca"]])[pc2_viz],3))
    x_label = paste0(paste0(paste0("PC",pc1_viz), ' stdev:  '),round(Stdev(object = seur_obj[["pca"]])[pc1_viz],3))
    PC_dimred <- DimPlot(seur_obj, reduction = "pca", dims = c(pc1_viz,pc2_viz))+labs(y= y_label, x = x_label)
    PC1_genes <- DimHeatmap(seur_obj, dims = c(pc1_viz), fast = FALSE)
    PC2_genes <- DimHeatmap(seur_obj, dims = c(pc2_viz), fast = FALSE)
    final_plot <- grid.arrange(as_grob(PC_dimred),as_grob(PC2_genes),as_grob(PC1_genes),
                                   ncol=2,
                                   as.table=TRUE)
                                   #heights=c(3,1))
    print(final_plot)}
  dev.off()
DimPlot(seur_obj, reduction = "pca")

```

umap regular normalization
```{r}
umap_col = "timepoint"
umap_col2 = "descriptive_name"
# PCs used for different UMAP representations
pcs_for_overview = c(2,3,4,5,6,7,8,9,10, 12, 15, 20)
# Generating a combined plot with only a legend in the first plotted (since this will be the same for the others)
plot.list <- list()
plot.list2 <- list()
label.vector <- c(umap_col, umap_col2)
j = 0

## This could be done nicer with a loop probably, in which for each principle component of interest, there is a umap run, and 2 different labels are shown for it.

for (i in (1:length(pcs_for_overview))){
  seur_obj <- RunUMAP(seur_obj, dims = 1:pcs_for_overview[i])
  dimnr <- as.character(pcs_for_overview[i])
  print(dimnr)
  if (i == 1){
    plot.list[[dimnr]] <- DimPlot(seur_obj, reduction = "umap", group.by = label.vector[1], combine = TRUE) + ggtitle(paste0("UMAP 1:", dimnr))
    plot.list2[[dimnr]] <- DimPlot(seur_obj, reduction = "umap", group.by = label.vector[2], combine = TRUE) + ggtitle(paste0("UMAP 1:", dimnr))
    i = i + 1
  }
  else {
    plot.list[[dimnr]] <- DimPlot(seur_obj, reduction = "umap", group.by = label.vector[1], combine = TRUE) + ggtitle(paste0("UMAP 1:", dimnr)) + theme(legend.position = "none")
    plot.list2[[dimnr]] <- DimPlot(seur_obj, reduction = "umap", group.by = label.vector[2], combine = TRUE) + ggtitle(paste0("UMAP 1:", dimnr)) + theme(legend.position = "none")
  i = i + 1
  }
}
pdf(paste(resultsdir, "6a.norm_UMAPdiffsettings.pdf", sep = '/'), width = 20, height = 15)
CombinePlots(plot.list, nrows = round(length(pcs_for_overview)/3))
print(ElbowPlot(seur_obj))
dev.off()

```

```{r}
seur_obj <- RunUMAP(seur_obj, dims = 1:8)
seur_obj <- FindNeighbors(seur_obj, dims = 1:8) 
seur_obj <- FindClusters(seur_obj, verbose = FALSE, resolution = 0.5, graph.name = "sf_snn")
seur_obj <- BuildClusterTree(seur_obj)
  
pdf(paste(resultsdir,'7a.norm_umap.pdf',sep="/") ,width=8,height=8,paper='special') 
#DimPlot(seur_obj, label = TRUE)
DimPlot(seur_obj, label=FALSE, group.by= "timepoint")+ ggtitle("timepoint") 
DimPlot(seur_obj, label=FALSE, group.by= "cell_type")+ ggtitle("original identity") 
DimPlot(seur_obj, label=TRUE, group.by = 'seurat_clusters') + ggtitle("Louvain Clustering") + ggtitle(paste0("cluster resolution ", 0.5))
PlotClusterTree(object = seur_obj)
DimPlot(seur_obj, label=FALSE, group.by= "batch")+ ggtitle("batch") 
DimPlot(seur_obj, label=FALSE, group.by= "descriptive_name")+ ggtitle("name") 
FeaturePlot(seur_obj, features = 'nCount_sf', cols =c('white','purple'))
FeaturePlot(seur_obj, features = 'nFeature_sf', cols =c('white','purple'))
FeaturePlot(seur_obj, features = 'altexps_MT_percent', cols =c('white','purple'))
DimPlot(seur_obj, group.by = 'Phase')
FeaturePlot(seur_obj, features = 'S.Score', cols =c('white','dodgerblue3'))
FeaturePlot(seur_obj, features = 'G2M.Score', cols =c('white','green4'))
dev.off()
```

#lets first perform clustering, after which we can start to quantify gene expression in all the clusters.
```{r}
pdf(paste(resultsdir, "8.clustering_resolution.pdf", sep = '/'), width = 20, height = 8)
for (i in 1:20){
  res <- i/10
  cluster_variable_name <- paste0("sf_snn_res.", res)
  seur_obj <- FindClusters(seur_obj, verbose = FALSE, resolution = res, graph.name = "sf_snn")
  seur_obj <- BuildClusterTree(seur_obj)
  p1 <- DimPlot(seur_obj, label=TRUE, group.by = 'timepoint') + ggtitle(paste0("cluster resolution ", res))
  p2 <- DimPlot(seur_obj, label=TRUE, group.by = 'seurat_clusters') + ggtitle("Louvain Clustering") + ggtitle(paste0("cluster resolution ", res))
  p3 <- ggplot(seur_obj@meta.data, aes(eval(parse(text= cluster_variable_name)), fill = timepoint))+geom_bar(stat="count")
  print(multiplot(p1, p2, p3, cols = 3))
  PlotClusterTree(object = seur_obj)


}

clustree(seur_obj, prefix = "sf_snn_res.", node_colour = 'sc3_stability')
clustree(seur_obj, prefix = "sf_snn_res.")

dev.off()


#0.55 seems the best cluster resolution to find heterogeneity that isńt cell cycle based
seur_obj <- FindClusters(seur_obj, verbose = FALSE, resolution = 0.50 , graph.name = "sf_snn")
DimPlot(seur_obj, label=TRUE, group.by = 'seurat_clusters') + ggtitle("Louvain Clustering") 
table(seur_obj$cell_type)
```

#lets merge cell cycle related clusters
```{r}
current.cluster.ids <- c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9)
new.cluster.ids <- c('ESC', 'd11_1', 'LSC', 'LSC', 'ABCG2_cells', 'multipotent_1', 'd24_prop_diff', 'd11_2', 'ESC', 'multipotent_2')
seur_obj$costum_clustering <- plyr::mapvalues(x = seur_obj$sf_snn_res.0.5, from = current.cluster.ids, to = new.cluster.ids)
cluster_order <- c('ESC','multipotent_1','multipotent_2','d11_1','d11_2', 'ABCG2_cells', 'd24_prop_diff', 'LSC')



# Re-level object@ident
seur_obj$costum_clustering <- factor(x = seur_obj$costum_clustering, levels = cluster_order)

pdf(paste(resultsdir, "8b.clustering_costum.pdf", sep = '/'), width = 20, height = 8)
p1 <- DimPlot(seur_obj, label=TRUE, group.by = 'timepoint')
p2 <- DimPlot(seur_obj, label=TRUE, group.by = 'costum_clustering') + ggtitle("Louvain Clustering")
p3 <- ggplot(seur_obj@meta.data, aes(costum_clustering, fill = timepoint))+geom_bar(stat="count")+ theme(axis.text.x = element_text(angle = 90))

print(multiplot(p1, p2, p3, cols = 3))
dev.off()

Idents(seur_obj) <- "costum_clustering"

# cluster re-assignment occurs, which re-assigns clustering in my_levels (assuming you have 12 clusters in total)
#my_levels <- c(7, 6, 4, 3, 1, 2, 5, 10, 8, 0, 9, 11, 12)

```

Lets find marker genes for each cluster
```{r}
cluster.markers <- FindAllMarkers(seur_obj, only.pos = TRUE)

#n_genes <- 40
#heatmap.markers <- cluster.markers %>% group_by(cluster) %>% top_n(n_genes, avg_logFC)
#DoHeatmap(seur_obj, features = heatmap.markers$gene) + NoLegend()

#filter on only sigi
cluster.markers_fc <- cluster.markers[cluster.markers$p_val_adj < 0.01,]

cluster.markers_fc <- cluster.markers_fc[(cluster.markers_fc$avg_log2FC > 0.58) ,]
table(cluster.markers_fc$cluster)
write.csv(cluster.markers, paste0(resultsdir,"/marker_genes.csv"))
```

#lets add the top10 of each cluster their marker genes to a file to vizualize
```{r}

cluster_markers = list()

for (cluster in unique(cluster.markers_fc$cluster)){
  print(cluster)
  deg_subset <- cluster.markers_fc[cluster.markers_fc$cluster == cluster,]
  cluster_genes <- head(deg_subset, 20)
  cluster_markers[cluster] = list(cluster_genes$gene)
}
cluster_markers = t(as.data.frame(cluster_markers))

write.table(cluster_markers, paste0(resultsdir,"/top_markers.csv"), col.names = F, sep =",")
```

```{r}
plot_list <- list()
cluster_GOs <- c() 
expressed_genes <- row.names(counts(sce_qc)[rowSums(counts(sce_qc))>20,])

for (cluster in unique(cluster.markers_fc$cluster)){
  print(cluster)
  df_subset <- cluster.markers_fc[cluster.markers_fc$cluster == cluster,]
  cluster_genes <- unique(str_sub(df_subset$gene,end = -1))
  PC1_ego <- enrichGO(gene = cluster_genes,
                          universe = expressed_genes,
                          OrgDb         = org.Hs.eg.db,
                          keyType       = "SYMBOL",
                          ont           = "BP",
                          pAdjustMethod = "BH",
                          #pvalueCutoff  = 0.01,
                          qvalueCutoff  = 0.05)
  if(is.null(PC1_ego)){
        go_plot1 <- rectGrob(gp=gpar(col=NA))
        }else{
        go_plot1 <- barplot(PC1_ego, showCategory=10)
        }
      plot_list <- c(plot_list, list(go_plot1))
      cluster_GOs <- append(cluster_GOs, (paste('cluster', cluster, sep = '')))
}
grob2 = ggarrange(plotlist = plot_list , ncol =1, labels = cluster_GOs) #align = 'v')



```

Heatmap cluster marking genes:
```{r}
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}

heatmap.markers <- cluster.markers_fc
cluster.markers_fc$log2FC <- cluster.markers_fc$avg_logFC 
top10 <- cluster.markers_fc %>%  group_by(cluster) %>%  top_n(n = 5)
top5_genes <- DoHeatmap(seur_obj, features = top10$gene) + NoLegend()
top10 <- cluster.markers_fc %>% group_by(cluster) %>% top_n(n = 10, wt = 'avg_log2FC')
marker_top5 <- DoHeatmap(seur_obj, features = top10$gene) + NoLegend()
n_genes <- 100
heatmap.markers <- cluster.markers_fc %>% group_by(cluster) %>% top_n(n_genes, avg_log2FC)

cell_type_plot <- DimPlot(seur_obj, label=TRUE, group.by = 'timepoint',pt.size = 2) + ggtitle("Timepoint") 
cluster_plot <- DimPlot(seur_obj, label=TRUE, group.by = 'costum_clustering', pt.size = 2) + ggtitle("Louvain Clustering") 
grob_clustering = ggarrange(plotlist = list(cell_type_plot, cluster_plot) , ncol =1) #labels = cluster_GOs) #align = 'v')

nclust<- length(unique(as.numeric(cluster.markers_fc$cluster)))

RGB_colours_ggplot <- as.list(as.character(gg_color_hue(nclust)))
names <- as.numeric(unique(heatmap.markers$cluster))
col_fun = colorRamp2(names, RGB_colours_ggplot)#make sure the rows correspond to ggplot collour mapping

mat <- as.matrix(seur_obj@assays$sf@scale.data[heatmap.markers$gene,])
mat <- rbind(mat,seur_obj@meta.data$costum_clustering)
mat <- mat[,order(mat[nrow(mat),])]

column_ha <- HeatmapAnnotation(cluster = mat[nrow(mat),], col =list(cluster = col_fun))
#column_ha <- HeatmapAnnotation(cluster = unique(heatmap.markers$cluster),], col =list(cluster = col_fun))

breaks <- mat[nrow(mat),] 
mat <- mat[-nrow(mat),] 

row_ha = rowAnnotation(adj_p_vallue = heatmap.markers$p_val_adj)

clust_heatmap <- grid.grabExpr(draw(Heatmap(mat, column_split = breaks,row_split = as.numeric(heatmap.markers$cluster), cluster_columns = F, cluster_rows = F,show_row_names = F,show_column_names = F,row_names_gp = gpar(fontsize = 6), top_annotation = column_ha, left_annotation = row_ha, row_names_rot = -40)))

top5_cluster_markers <- cluster.markers_fc %>%  group_by(cluster) %>%  top_n(n = 5)
# df_wider <- pivot_wider(df, names_from = cluster, values_from = gene_name_shorter) 
# separate_rows(df_wider,sep = "[^[:alnum:].]+", convert = FALSE)
df <- top5_cluster_markers
pdf(paste0(resultsdir,'/9.Complex_heatmap.pdf') ,width=40,height=16,paper='special') 
plot_grid(clust_heatmap,grob2,grob_clustering,ncol = 3, rel_heights = c(1, 4,2), rel_widths = c(3,2,2))
plot.new()
grid.table(df)
dev.off()
```


#save environement to run velocity upon
```{r}
saveRDS(seur_obj, file = paste0(resultsdir,"/seur_obj.rds"))
seur_obj <- readRDS(paste0(resultsdir,"/seur_obj.rds"))
```

#run progeny 
```{r}
CellsClusters <- data.frame(Cell = names(Idents(seur_obj)), 
    CellType = as.character(Idents(seur_obj)),
    stringsAsFactors = FALSE)

seur_obj@assays$RNA <- seur_obj@assays$sf 
seur_obj <- progeny(seur_obj, scale=FALSE, organism="Human", top=500, perm=1,  return_assay = TRUE)

## We can now directly apply Seurat functions in our Progeny scores. 
## For instance, we scale the pathway activity scores. 
seur_obj <- Seurat::ScaleData(seur_obj, assay = "progeny") 

progeny_scores_df <- 
    as.data.frame(t(GetAssayData(seur_obj, slot = "scale.data", 
        assay = "progeny"))) %>%
    rownames_to_column("Cell") %>%
    gather(Pathway, Activity, -Cell) 

progeny_scores_df <- inner_join(progeny_scores_df, CellsClusters)

summarized_progeny_scores <- progeny_scores_df %>% 
    group_by(Pathway, CellType) %>%
    summarise(avg = mean(Activity), std = sd(Activity))

summarized_progeny_scores_df <- summarized_progeny_scores %>%
    dplyr::select(-std) %>%   
    spread(Pathway, avg) %>%
    data.frame(row.names = 1, check.names = FALSE, stringsAsFactors = FALSE) 

paletteLength = 100
myColor = colorRampPalette(c("Darkblue", "white","red"))(paletteLength)

progenyBreaks = c(seq(min(summarized_progeny_scores_df), 0, 
                      length.out=ceiling(paletteLength/2) + 1),
                  seq(max(summarized_progeny_scores_df)/paletteLength, 
                      max(summarized_progeny_scores_df), 
                      length.out=floor(paletteLength/2)))


mat <- t(summarized_progeny_scores_df[,-1])


pathway_viz_object <- seur_obj
pathways <- rownames(pathway_viz_object@assays$progeny@scale.data)
pathways <- paste("progeny_targets_", pathways, sep="")
rownames(pathway_viz_object@assays$progeny@scale.data) <- pathways
DefaultAssay(pathway_viz_object) <- "progeny"

#VlnPlot(pathway_viz_object, combine=F, features = pathways, group.by = 'costum_clustering', assay = 'progeny', slot = 'scale.data')

pdf(paste0(resultsdir,'/10.progeny_pwahtway_targets_activation.pdf') ,width=10,height=10,paper='special') 
Heatmap(mat, cluster_columns =T)
FeaturePlot(pathway_viz_object, features = pathways, pt.size = 0.2, ncol = 4, combine=F, slot = 'scale.data', blend.threshold = 4)

dev.off()

```

Lets vizualize both the top markers from each cluster, and the markers from the marker gexpression file
```{r}
marker_gene_file <- "/ceph/rimlsfnwi/data/moldevbio/zhou/jsmits/scRNAseq_iPSC_iLSCs/LSC_Marker_Genes.csv"
top_cluster_file <- paste0(resultsdir,"/top_markers.csv")

marker_dir <- paste0(resultsdir, '/marker_genes/')
print(marker_dir)
system(paste("mkdir -p ", marker_dir))


for (genefile in c(marker_gene_file, top_cluster_file)){
  marker_genes_df <- read.table(genefile, row.names=1, header = FALSE, sep = ',',comment.char = "#", stringsAsFactors = F)
  for (row in 1:nrow(marker_genes_df)) {
  
      filename <- rownames(marker_genes_df[row,])
      filename <- paste0(marker_dir,filename)
      marker_genes <- as.list(marker_genes_df[row,])
      marker_genes <- as.list(marker_genes[ marker_genes != ""])
  
      n <- length(marker_genes)
      k <- 10 ## your LEN
      marker_gene_sets <- split(marker_genes, rep(1:ceiling(n/k), each=k)[1:n])
  
      pdf(paste0(filename,'.pdf'), width = 20, height = 15)
  
      for (genes in  marker_gene_sets){#loop over subsets of max 12 genes to vizualize
        plot_list1 <- DimPlot(seur_obj, label=TRUE, group.by = 'timepoint', combine = F, pt.size = 0.1)
        plot_list1 <- append(plot_list1, DimPlot(seur_obj, label=TRUE, group.by = 'costum_clustering', combine = F, pt.size = 0.1))
        count_plots <- plot_list1
        count_plots_clusters <- plot_list1
        count_plots <- append(count_plots, try(VlnPlot(seur_obj, combine=F, features = as.vector(unlist(genes)), ncol = 4, group.by = 'timepoint', assay = 'sf')))
        count_plots_clusters <- append(count_plots_clusters, try(VlnPlot(seur_obj, combine=F, features = as.vector(unlist(genes)), ncol = 4, group.by = 'costum_clustering', assay = 'sf')))
  
      plot_list1 <- append(plot_list1,FeaturePlot(seur_obj, features = as.vector(unlist(genes)), pt.size = 0.2, ncol = 4, combine=F))
      print(cowplot::plot_grid(plotlist = plot_list1))
      #print(FeaturePlot(seur_obj, features = genes, pt.size = 0.2, ncol = 4))

      p1 <- cowplot::plot_grid(plotlist = count_plots)
      title <- ggdraw() + draw_label("seurat norm int of marker genes per timepoint", fontface = 'bold')
      print(cowplot::plot_grid(title, p1, ncol = 1, rel_heights = c(0.1, 1)))
      p1 <- cowplot::plot_grid(plotlist = count_plots_clusters)
      title <- ggdraw() + draw_label("seurat norm int of marker genes per timepoint", fontface = 'bold')
      print(cowplot::plot_grid(title, p1, ncol = 1, rel_heights = c(0.1, 1)))
  
      }
    dev.off()
      
      
      
    }}


```

Export the seurat object for shiny app vizualization
```{r}
DimPlot(seur_obj, reduction = "umap")

# make the SingleCellExperiment object from the Seurat object:
sce_shiny <- as.SingleCellExperiment(seur_obj)
# check if coordinates were transferred well:
saveRDS(sce_shiny,paste0(resultsdir,'/seur_obj_shiny.rds'))
```

#run monocle pseudotime
```{r}
library('monocle3')
seur_obj <- readRDS('/ceph/rimlsfnwi/data/moldevbio/zhou/jsmits/scRNAseq_iPSC_iLSCs/analysis/20210721/seur_obj.rds')
#cds <- updateCDS(seur_obj)
cds <- as.cell_data_set(seur_obj)
cds <- estimate_size_factors(cds)
cds = cluster_cells(cds, reduction_method = "UMAP") #Monocle3 Pipeline
cds = learn_graph(cds, use_partition = F, close_loop = F) #Change Partition and or Close_loop as desired
cds = order_cells(cds)
pdf(paste0(resultsdir,'/13.monocle3_pseudotime.pdf') ,width=10,height=10,paper='special') 
plot_cells(
  cds = cds,
  color_cells_by = "pseudotime",
  show_trajectory_graph = TRUE, cell_size = 2
) + ggtitle("UMAP of Pseudotime")
dev.off()
#plot_cells(cds, genes=c("PAX6", "KRT14"), cell_size = 2)
# 
# #Subset ad and hc, perform monocle3 pipeline
# interad <- inter[,grepl("AD", colData(inter)$disease, ignore.case=TRUE)]
# interad = preprocess_cds(interad)
# pr_graph_test_res <- graph_test(interad,neighbor_graph="knn" )
# pr_deg_ids <- row.names(subset(pr_graph_test_res, q_value < 0.05))#maybe p val?
# gene_module_df <- find_gene_modules(interad[pr_deg_ids,], resolution=1e-2)
# sig_gene_namesad <- row.names(interad[gene_module_df$id])
# 
# interhc <- inter[,grepl("HC", colData(inter)$disease, ignore.case=TRUE)]
# interhc = preprocess_cds(interhc)
# pr_graph_test_res <- graph_test(interhc)
# pr_deg_ids <- row.names(subset(pr_graph_test_res, q_value < 0.05))
# gene_module_df <- find_gene_modules(interhc[pr_deg_ids,], resolution=1e-2)
# sig_gene_nameshc <- row.names(interhc[gene_module_df$id])

```


#lets try to integrate the data with the LAKO scCornea atlas
```{r}

#seur_obj <- readRDS(paste0(resultsdir,"/seur_obj.rds"))
seur_obj <- readRDS('/ceph/rimlsfnwi/data/moldevbio/zhou/jsmits/scRNAseq_iPSC_iLSCs/analysis/20210721/seur_obj.rds')
LAKO_obj <- readRDS('/ceph/rimlsfnwi/data/moldevbio/zhou/jarts/R/scRNA-seq/lakoRNAannotated.rds')

reference <- LAKO_obj
reference$id <- 'reference'
seur_obj$id <- 'query'

pdf(paste0(resultsdir,'/11.objects_before_merge.pdf') ,width=10,height=10,paper='special') 
DimPlot(seur_obj, label=TRUE, group.by = 'costum_clustering')
DimPlot(reference, label=TRUE, group.by = 'costum_clustering')
dev.off()
```

```{r}
#https://satijalab.org/seurat/articles/integration_introduction.html

obj_list <- c(seur_obj, reference)
obj_list <- lapply(X = obj_list, FUN = function(x) {
    x <- NormalizeData(x)
    x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 1000)
})
# select features that are repeatedly variable across datasets for integration
features <- SelectIntegrationFeatures(object.list = obj_list)

#epithelial anchors:
epi.anchors <- FindIntegrationAnchors(object.list = obj_list, anchor.features = features)

#create integrated data assay
epi_combined <- IntegrateData(anchorset = epi.anchors)

#
DefaultAssay(epi_combined) <- "integrated"

# Run the standard workflow for visualization and clustering
epi_combined <- ScaleData(epi_combined, verbose = FALSE)
epi_combined <- RunPCA(epi_combined, npcs = 40, verbose = FALSE)
epi_combined <- RunUMAP(epi_combined, reduction = "pca", dims = 1:20)
epi_combined <- FindNeighbors(epi_combined, reduction = "pca", dims = 1:20)
epi_combined <- FindClusters(epi_combined, resolution = 0.15)

#viz
pdf(paste0(resultsdir,'/12.CCA_merge.pdf') ,width=10,height=10,paper='special') 
DimPlot(epi_combined, reduction = "umap", group.by = "id")
DimPlot(epi_combined, reduction = "umap", group.by = 'costum_clustering',label = TRUE, repel = TRUE)
DimPlot(epi_combined, reduction = "umap", group.by = "seurat_clusters")
epi_combined <- RenameIdents(epi_combined, `0` = "Corneal Basal", `1` = "Conjunctival", `2` = "Corneal_stromal",
    `3` = "LSCs", `4` = "multipotent", `5` = "LNPCs", `6` = "stromal", `7` = "d11", `8` = "VES", `9` = "ABCG2",
    `10` = "FCECs", `11` = "MEC", `12` = "IC", `13` = "bad_cells", `14` = "bad_cells", `15` = "bad_cells")
DimPlot(epi_combined, reduction = "umap",label = TRUE, repel = TRUE)

dev.off()

DefaultAssay(epi_combined) <- "RNA"
LSC_markers <- FindConservedMarkers(epi_combined, ident.1 = 'LSC', grouping.var = "id", verbose = FALSE)
head(LSC_markers)




```





```{r}
#Fast integration using reciprocal PCA (RPCA)

obj_list <- lapply(X = obj_list, FUN = function(x) {
    x <- ScaleData(x, features = features, verbose = FALSE)
    x <- RunPCA(x, features = features, verbose = FALSE)
})

epi_anchors_RPCA <- FindIntegrationAnchors(object.list = obj_list, anchor.features = features, reduction = "rpca")


#create integrated data assay
epi_combined <- IntegrateData(anchorset = epi_anchors_RPCA)


DefaultAssay(epi_combined) <- "integrated"

# Run the standard workflow for visualization and clustering
epi_combined <- ScaleData(epi_combined, verbose = FALSE)
epi_combined <- RunPCA(epi_combined, npcs = 40, verbose = FALSE)
epi_combined <- RunUMAP(epi_combined, reduction = "pca", dims = 1:20)
epi_combined <- FindNeighbors(epi_combined, reduction = "pca", dims = 1:20)
#epi_combined <- FindClusters(epi_combined, resolution = 0.5)

#viz
pdf(paste0(resultsdir,'/12.RPCA_merge.pdf') ,width=10,height=10,paper='special') 
DimPlot(epi_combined, reduction = "umap", group.by = "id")
DimPlot(epi_combined, reduction = "umap", group.by = 'costum_clustering',label = TRUE, repel = TRUE)
dev.off()
```









```{r}
anchors <- FindTransferAnchors(
  reference = LAKO_obj,
  query = seur_obj,
  reference.reduction = "pca",
  dims = 1:10
)
```

```{r}
iLSC_on_LAKO <- MapQuery(
  anchorset = anchors,
  query = seur_obj,
  reference = LAKO_obj,
  refdata = list(
    costum_clustering = "costum_clustering",
    predicted_ADT = "ADT"
  ),
  reference.reduction = "pca", 
  reduction.model = "umap"
)
```




```{r}
ifnb.list <- list(seur_obj, LAKO_obj)

features <- SelectIntegrationFeatures(object.list = ifnb.list)

immune.anchors <- FindIntegrationAnchors(object.list = ifnb.list, anchor.features = features)
immune.combined <- IntegrateData(anchorset = immune.anchors)
```

```{r}
# specify that we will perform downstream analysis on the corrected data note that the
# original unmodified data still resides in the 'RNA' assay
DefaultAssay(immune.combined) <- "integrated"

# Run the standard workflow for visualization and clustering
immune.combined <- ScaleData(immune.combined, verbose = FALSE)
immune.combined <- RunPCA(immune.combined, npcs = 30, verbose = FALSE)
immune.combined <- RunUMAP(immune.combined, reduction = "pca", dims = 1:30)
immune.combined <- FindNeighbors(immune.combined, reduction = "pca", dims = 1:30)
immune.combined <- FindClusters(immune.combined, resolution = 0.5)
```


```{r}
DimPlot(immune.combined, reduction = "umap", group.by = 'costum_clustering',label = TRUE, repel = TRUE)
DimPlot(immune.combined, reduction = "umap", group.by = 'orig.ident')

```
























